Analyzing Student Use of Spacing and Interleaving Strategies in Interactions with GenAI-Powered Chatbots in Programming Courses

### 2024-10-29 CSEDU2025/Paper.tex
update paper

\usepackage{epsfig}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{algorithm2e}
\usepackage[bottom]{footmisc}
% Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
\usepackage{natbib}
\usepackage{multirow}
\usepackage{SCITEPRESS}

@@

\title{
  Examining Programming Students’ Interaction with Generative AI: A
  Metacognitive Perspective
}

@@

\keywords{
  Programming Education, Generative Artificial Intelligence, Chatbots and
  Metacognition
}

\abstract{
This study examines the interactions of programming students with a chatbot,
CharlieBot, powered by OpenAI's GPT-3.5 and enhanced with the
Retrieval-Augmented Generation (RAG) technique, within the context of a Java
programming course. It investigates how students use the bot while employing two
metacognitive strategies: interleaving and spacing. Using the logs of student
interactions with the bot, the methodology involved classifying and analyzing
student prompts into eight distinct categories. The findings reveal that while
most interactions emphasize active learning strategies, students demonstrate
limited conscious application of interleaving and spacing. This study
underscores the need for more guidance on the strategic use of AI
tools to improve learning outcomes.
}

@@

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} indicate that more than 54,5\% of students are already using
GenAI for homework, likely due to their perception of this technology as
advantageous and intuitive \citep{Boudouaia24}. All of these studies highlight
the need to increase the understanding how students interact with these tools
and how they can be used to improve learning, as emphasized \cite{Lo24}.

In response to the growing need for deeper insight into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational bot powered by gpt-3.5 from OpenAI, enhanced with the
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of the metacognitive strategies
employed by students. To achieve this objective, we formulate three research
questions:

\begin{itemize}
  \item RQ1 - What was the distribution pattern of the classified messages?
  \item RQ2 - Was there a space between the student prompts? If so, which
  category promotes the largest spacing between submissions?
  \item RQ3 - Do students' conversations with the bot alternate between
  practical and theoretical approaches, thereby promoting interleaving?
\end{itemize}

@@

This paper is organized as follows: Section 2 presents the theoretical
framework, Section 3 describes the material and methods used in the study,
Section 4 presents the results and discussion, and Section 5 concludes the
paper and outlines future work.

@@

This section presents the theoretical framework that underpins the study. First,
we introduce the concept of metacognition used in this study. Next, we discuss
related work that explores the use of metacognitive strategies and scaffolds
in educational settings.

@@

Given that generative AI often produces automatic and highly accurate responses
\citep{Puryear22}, it is essential for students to use these tools within an
active learning process to enhance their learning outcomes. This active
learning process can be further understood through the lens of Metacognition,
which focuses on the awareness of one’s mental processes. \cite{flavell79}
proposed a model of metacognitive monitoring that includes four interrelated
phenomena: Metacognitive Knowledge, Metacognitive Experience, Metacognitive
Goals, and Metacognitive Actions. These processes do not occur in isolation, but
they influence each other, altering cognitive progress over time.

@@

Spacing and Interleaving are two metacognitive strategies that have been shown
to enhance learning outcomes and supports the research questions of this study.
Spacing refers to the practice of distributing study sessions over time, which
has been shown to improve long-term retention and understanding of the material
\citep{Carvalho20}. Interleaving involve mixing different topics or problems
within a single learning session, which has been shown to also enhance long-term
learning and the application of student's knowledge in other contexts and
situations \citep{Rivers21}.

Since students are often dealing with a single problem during an interaction
session with the bot, in this study we propose a specialized form of the
Interleaving strategy, termed the \textit{Perspective Interleaving Hypothesis}.
Unlike traditional Interleaving, which typically involves alternating between
different topics or problem types, Perspective Interleaving can be characterized
by a metacognitive strategy that intentionally switches between angles of a
study object.

An important factor supporting the hypothesis is that when switching between
categories, such as an abstraction and a concrete implementation, a student is
engaging in Discriminative Contrasts \citep{kang12}. Furthermore, the hypothesis
that this approach enhances student learning may be grounded on Kolb’s
Experiential Learning Theory, which posits that learning involves a continuous
and structured cycle of concrete experiences, reflective observation, abstract
conceptualization, and active experimentation \citep{kolb14}.

@@

\cite{Margulieux24} conducted a study on how undergraduate students in
introductory programming courses used generative AI to solve programming
problems in a naturalistic setting. The research focused on examining the
relationship between AI usage and students’ self-regulation strategies,
self-efficacy, and fear of failure in programming. Additionally, the study
explored how these variables interacted with learner characteristics, perceived
usefulness of AI, and academic performance. Findings revealed that students with
higher self-efficacy, lower fear of failure, or higher prior grades tended to
use AI less frequently or  later in the problem-solving process and perceived it
as less useful compared to their peers. However, no significant relationship was
found between students’ self-regulation strategies and their use of AI.

@@

\cite{Sidra24} addressed the growing importance of metacognitive skills in an
AI-enabled workforce. The problem discussed was how AI systems, especially
collaborative tools like large language models (LLMs), impacted human
decision-making and the challenges they posed due to the lack of human-like
metacognition in AI. The method used identified four main characteristics that
differentiated human-AI interaction from human-human collaboration and proposed,
based on the dual-process theory of human thought, that metacognitive thinking
was crucial for the effective use of AI. The results indicated that human
workers needed to develop and apply metacognitive skills—such as planning,
monitoring, and evaluating their thought processes—to mitigate cognitive biases
and improve decision-making when working with AI. The study concluded by
recommending training and AI design improvements to enhance workers’
metacognitive abilities, ensuring more effective collaboration with AI systems.

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing
improvements in problem-solving and collaboration. \cite{Khusnul24} investigates
the efficacy of a meta-learning approach in improving metacognitive and creative
skills. This study confirms the effectiveness of meta-learning strategies and
elucidates the relationship between meta-learning and metacognition. This
research suggests that meta-learning can improve metacognitive abilities,
providing valuable insights into educational technology and course design in
higher education settings.

@@

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis. Initially, \cite{Ghimire24}
proposed four categories: Debugging Help (DH), Conceptual Question (CQ), Code
Snippet (CS), and Complete Solution (CSO). However, the data collected indicated
the need for additional categories, leading to the inclusion of four more:
Multiple Question (MQ), Student Corrections (SC), Language Change (LC), and
Uncategorized (U). In this study, students were free to use the bot as they
wanted, which contributed to the emergence of these additional categories.
Table \ref{tab:categories} presents these categories along with their respective
descriptions.

@@

@@

A total of 343 student messages were categorized in 42 conversations, with
an average of 8.17 messages per conversation. This data offers insights into how
users engage with CharlieBot and the types of prompts they submit. The results of
the analysis are presented in this section, addressing the research questions
outlined in the introduction.

Figure \ref{fig:graph1} shows the distribution of the messages in the
categories:

\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.62]{img/figure1.png}
    \caption{Classification of messages.}
    \label{fig:graph1}
\end{figure}

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.52]{img/figure2.png}
  \caption{Examples of conversations.}
  \label{fig:graph2}
\end{figure*}

\subsection{Classification of Messages}

About the first research question: \textit{RQ1: What was the distribution
pattern of the classified messages?}

According to Figure \ref{fig:graph1} messages classified as
Conceptual Question account for 32.1\% of the total. Often, the sequence of
Conceptual Question begins with a request for a Complete Solution or Code
Snippet, which then evolves into a series of Conceptual Question, as seen in
conversation 2 of Figure \ref{fig:graph2}, which illustrates some examples of
students' conversations with the bot. In other cases, the conversation
consists entirely of messages classified as Conceptual Question, such as in
conversation 10 of Figure \ref{fig:graph2}. This type of message indicates that
students recognize that they have not yet fully mastered a concept and are
seeking to better understand it. In this context, the student realizes that
solving the problem requires understanding of concepts, thus applying an active
learning strategy.

As shown in Figure \ref{fig:graph1}, approximately 30.9\% of the messages were
classified as Debugging Help. These prompts often begin with a request for a
Complete Solution or Code Snippet. Following this, students typically switch from
passive to more active learning strategy, as illustrated by the start
of conversation 12 in Figure \ref{fig:graph2}. Debugging Help messages represent
a form of active learning, in which students likely focus on gaining a practical
understanding of the code, identifying errors, and improving their debugging
skills for future tasks.

Since most of the messages are located in categories where students are
notoriously engaged in being more active in their learning (Conceptual Question
and Debugging Help), corroborating the findings of \cite{Ghimire24} which
showed that the questions are also localized in the same categories.

Requests for the bot to generate a Code Snippet or Complete Solution accounted
for 10.5\% of student messages. These prompts reflected a desire for more
direct answers; however, it was observed that, after receiving these solutions,
many students transitioned to a more active approach in seeking to understand
the code or the underlying concepts. Thus, it can be inferred that the prompts
within the Code Snippet or Complete Solution category were often used as a
starting point for a more in-depth study.

Multiple Choice exercise resolutions represented 16.9\% of the responses.
These prompts typically involved students submitting one or more exercises to
CharlieBot and requesting solutions. We identified sequences of multiple choice
questions, as illustrated by conversations 8, 9, and 15 in \ref{fig:graph2}.
This less active learning approach can occur for various reasons, ranging from
the belief that reviewing solutions to similar questions enhances learning, to
issues such as negligence, poor planning, or other possible factors.

Student corrections to bot responses accounted for 2.6\% of the
interactions. In most cases, these corrections occurred when the student already
had the correct answer to an exercise and identified an error in the bot
response, as illustrated in conversations 4 and 9 in Figure \ref{fig:graph2}.
These corrections highlight that, although the bot provides quick feedback, it
is not always accurate and its errors can be difficult for students to
identify.

Figure \ref{fig:graph1} also shows that around 5\% of the prompts written by
students were not categorized. These messages range from greetings to the Bot,
such as a simple "\textit{Hello}," to contextual statements like
"\textit{I'm reviewing object orientation}"

@@

% Spacing

About the second research question: \textit{RQ2: Was there a space between the
student prompts? If so, which category promotes the largest spacing between
submissions?}

Approximately 26.2\% of the conversations, or roughly a quarter, included pauses
longer than 60 minutes. However, this data only indicates that students paused
their use of the AI, which could suggest that they were either reflecting on the
topic independently or engaging in unrelated activities. Furthermore, 16.7\%
of the conversations exhibited pauses of 12 hours and 9\% showed delays
exceeding 24 hours between messages, indicating that some degree of spacing
occurred. However, it is important to note that this behavior may have been
influenced by the design of the initial version of CharlieBot, which only kept
the chat session active as long as the student kept the browser tab open and
avoided reloading the page. This adds uncertainty about the precise context in
which students applied spacing.

There is a noticeable difference in the interaction time with the bot when a
student asks a conceptual question. The time gap creates an effect of
information retrieval when resuming the activity. Therefore, we can consider
this category as one in which students are more cognitively engaged. Thus, in
this category, students are likely more focused on understanding, which aligns
with the metacognitive goal of promoting comprehension and long-term retention.
Figure \ref{fig:graph3} shows the average interaction time per category.

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.55]{img/figure3.png}
  \caption{Average interaction time per category.}
  \label{fig:graph3}
\end{figure*}

\subsection{Perspective Interleaving}

% Interleaving

Regarding the third research question: \textit{RQ3: Do students' conversations
with the bot alternate between practical and theoretical approaches, thereby
promoting interleaving?}

When analyzing the classification of messages in conversations, we identified
the presence of sequences of messages belonging to the same category, which
could suggest a low occurrence of interleaving. To illustrate the presence of
blocks, Figure \ref{fig:graph2} presents examples of conversations and their
classification. In Conversation 2, the student alternates messages among Code
Snippet, Debugging Help, and Conceptual Question, showing a tendency to
interleave from a practical to theoretical perspective. In contrast, in
Conversation 10, the student focused on a single block of Conceptual Question.
Although the focus was on understanding a theoretical concept, this conversation
tended not to promote interleaving, due to the absence of alternation between
different types of task.

To better assess the extent of these sequences, we calculated the percentage of
messages that were part of repeated blocks. Considering blocks with at least
three messages, we observed that 56.2\% of the student's prompts were grouped
into some sequence. When analyzing blocks with a minimum of four messages, we
found that 47.5\% of the prompts were part of continuous sequences.

However, if we consider that sequential messages from a certain point indicate
that the student is \textit{blocked} (or focused) on a task, the remaining space
for interleaving is 43.8\%, assuming a block starting from at least three
consecutive messages, and 52.5\% for blocks starting from four consecutive
messages. Thus, the data suggest that interleaving occurs on students'
conversations with CharlieBot, but in which situations and with what type of
student profile is a study that needs to be further investigated.

@@

The results suggest that most student interactions with GenAI tools are centered
on active learning strategies, although the deliberate use of interleaving and
spacing techniques remains limited. Guiding students toward a more intentional
application of these strategies could enhance their learning outcomes. The
anonymity of interactions leaves it unclear whether these strategies were used
deliberately or emerged naturally, highlighting the need for future research on
students' conscious application of these methods. To support educators in
analyzing student behavior and offering more targeted assistance, an AI model
is being developed to automatically classify student messages, providing
valuable insights into how these tools are being utilized.

% TODO: Work in progress.

% * The results indicate that most interactions are focused on active learning
% strategies, although deliberate use of interleaving and spacing is limited. We
% believe that students can be guided \cite{Zheng19} \cite{Wang23} \cite{LiWei23}
% \cite{Khusnul24} to adopt a more deliberate approach in utilizing GenAI,
% enabling them to more effectively harness these tools to enhance their learning.

% * Due to the anonymity of the interactions, it's unclear whether
% students were intentionally using interleaving and spacing strategies or if
% these patterns emerged naturally. Future research should investigate whether
% students were consciously applying.

% * To provide teachers with valuable data for analyzing how students are using
% the tools, we are developing an AI model to automatically classify student
% messages. This will enable educators to better understand student behavior and
% offer more targeted support where needed.

\section*{ACKNOWLEDGEMENTS}

The authors would like to thank the Federal Institute of Education, Science and
Technology (IFRS) for the partial financial support provided for the execution
of this research.

@@

\section*{\uppercase{\textit{Appendix}}}

If any, the appendix should appear directly after the
references without numbering, and not on a new page. To do so please use the
following command: \textit{$\backslash$section*\{APPENDIX\}}

>>>>

\usepackage{epsfig}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{algorithm2e}
\usepackage[bottom]{footmisc}
% Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
\usepackage{natbib}
\usepackage{multirow}
\usepackage{SCITEPRESS}
\usepackage{arydshln}

@@

\title{
  Analyzing Student Use of Spacing and Interleaving Strategies in Interactions
  with GenAI-Powered Chatbots in Programming Courses
}

@@

\keywords{
  Programming, Generative Artificial Intelligence, Chatbots, Spacing,
  Interleaving and Metacognition
}

\abstract{
This study examines the interactions of programming students with a chatbot
powered by OpenAI's GPT-3.5 and enhanced with the Retrieval-Augmented Generation
(RAG) technique, within the context of a Java programming course. It
investigates how students use the bot while employing two metacognitive
strategies: interleaving and spacing. Using the logs of student interactions
with the bot, the methodology involved classifying and analyzing student prompts
into eight distinct categories. The findings reveal that while most interactions
emphasize active learning strategies, students demonstrate limited conscious
application of interleaving and spacing. This study underscores the need for
more guidance on the strategic use of AI tools to improve learning outcomes.
}

@@

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} indicate that more than 54,5\% of students are already using
GenAI for homework, likely due to their perception of this technology as
advantageous and intuitive \citep{Boudouaia24}. All of these studies highlight
the need to increase the understanding of how students interact with these tools
and how they can be used to improve learning, as emphasized \cite{Lo24}.

In response to the growing need for deeper insight into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational bot powered by OpenAI gpt-3.5, enhanced with the
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of two metacognitive strategies:
spacing and interleaving. To achieve this objective, we formulate three research
questions:

\begin{itemize}
  \item RQ1 - What distribution patterns emerged in the classified student
  interaction prompts with the educational chatbot?
  \item RQ2 - Was there spacing between student prompts, and if so, which
  category led to the most rapid and consistent interactions with the bot?
  \item RQ3 - Do students' interactions with the bot alternate between study
  topics to suggest interleaving?
\end{itemize}

@@

This paper is organized as follows. Section 2 presents the theoretical
framework, Section 3 describes the material and methods used in the study,
Section 4 presents the results and discussion, and Section 5 concludes the
paper and outlines future work.

@@

This section presents the theoretical framework that underpins the study. First,
we introduce the concept of metacognition used in this study. Next, we discuss
related work exploring the use of metacognitive strategies and scaffolds
in educational settings.

@@

Given that generative AI often produces highly accurate and automatic responses
\citep{Puryear22}, it is essential that students use these tools within an
active learning process to enhance their learning outcomes. This active
learning process can be further understood through the lens of metacognition,
which focuses on awareness of one's mental processes. \cite{flavell79}
proposed a model of metacognitive monitoring that includes four interrelated
phenomena: Metacognitive Knowledge, Metacognitive Experience, Metacognitive
Goals, and Metacognitive Actions. These processes do not occur in isolation, but
they influence each other, altering cognitive progress over time.

@@

Spacing and interleaving are two metacognitive strategies that have been shown
to enhance learning outcomes and support the research questions of this study.
Spacing refers to the practice of distributing study sessions over time, which
has been shown to improve long-term retention and understanding of the material
\citep{Carvalho20}. Interleaving involves mixing different topics or problems
within a single learning session, which has been shown to also enhance long-term
learning and the application of student's knowledge in other contexts and
situations \citep{Rivers21}.

@@

\cite{Margulieux24} conducted a study on how undergraduate students in
introductory programming courses used generative AI to solve programming
problems in a naturalistic setting. The research focused on examining the
relationship between AI usage and students’ self-regulation strategies,
self-efficacy, and fear of failure in programming. Furthermore, the study
explored how these variables interacted with the characteristics of the learners,
the perceived usefulness of AI, and academic performance. The findings revealed
that students with higher self-efficacy, lower fear of failure or higher prior
grades tended to use AI less frequently or later in the problem solving process
and perceived it as less useful compared to their peers. However, no significant
relationship was found between students’ self-regulation strategies and their
use of AI.

@@

\cite{Sidra24} addressed the growing importance of metacognitive skills in an
AI-enabled workforce. The problem discussed was how AI systems, especially
collaborative tools like large language models (LLMs), impacted human
decision-making and the challenges they posed due to the lack of human-like
metacognition in AI. The method used identified four main characteristics that
differentiated human-AI interaction from human-human collaboration and proposed,
based on the dual-process theory of human thought, that metacognitive thinking
was crucial for the effective use of AI. The results indicated that human
workers needed to develop and apply metacognitive skills, such as planning,
monitoring, and evaluating their thought processes, to mitigate cognitive biases
and improve decision-making when working with AI. The study concluded by
recommending training and AI design improvements to enhance workers’
metacognitive abilities, ensuring more effective collaboration with AI systems.

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing
improvements in problem-solving and collaboration. \cite{Khusnul24} investigates
the efficacy of a meta-learning approach in improving metacognitive and creative
skills. This study confirms the effectiveness of meta-learning strategies and
elucidates the relationship between meta-learning and metacognition. This
research suggests that metalearning can improve metacognitive abilities,
providing valuable insights into educational technology and course design in
higher education settings.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.62]{img/figure1.png}
  \caption{Categorization of messages.}
  \label{fig:graph1}
\end{figure}

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.45]{img/figure2.png}
  \caption{Examples of conversations.}
  \label{fig:graph2}
\end{figure*}

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis, representing different types of
prompt that students use when interacting with the bot. Initially,
\cite{Ghimire24} proposed four categories: Debugging Help (DH), Conceptual
Question (CQ), Code Snippet (CS), and Complete Solution (CSO). However, the data
collected indicated the need for additional categories, leading to the inclusion
of four more: Multiple Question (MQ), Student Corrections (SC), Language Change
(LC), and Uncategorized (U). In this study, students were free to use the bot as
they wanted, which contributed to the emergence of these additional categories.
Table \ref{tab:categories} presents these categories along with their respective
descriptions. To facilitate the analysis of interleaving, the prompt topics were
also categorized according to the course subjects, which are: (1) Java, (2)
Object Orientation, (3) Testing, (4) Recursion, (5) Data Structures, and (6)
Sorting and Searching Algorithms.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.39]{img/figure3.png}
  \caption{Average interaction time per category.}
  \label{fig:graph3}
\end{figure*}

@@

A total of 625 student messages were categorized in 81 conversations, with
an average of 7.72 messages per conversation. This data offers insights into how
users engage with CharlieBot and the types of prompts they submit. The results
of the analysis are presented in this section, addressing the research questions
outlined in the introduction. Figure \ref{fig:graph1} shows the distribution of
the messages in the categories.

\subsection{Categorization of messages}

About the first research question: \textit{RQ1: What distribution patterns
emerged in the classified student interaction prompts with the educational
chatbot?}

According to Figure \ref{fig:graph1} messages classified as
Conceptual Question account for 35.8\% of the total. Sequences of Conceptual
Questions often start from a practical perspective of a given code, as
illustrated in conversation 4 of Figure \ref{fig:graph2}, which presents
examples of students' interactions with the bot. In other cases, the
conversation consists entirely of messages classified as Conceptual Question,
such as in conversation 15 of Figure \ref{fig:graph2}.

As shown in Figure \ref{fig:graph1}, approximately 29.9\% of the messages were
classified as Debugging Help. Debugging help messages are likely to provide
students with a practical approach to understanding code, identifying errors,
and building skills to enhance their debugging abilities for future tasks.

Most of the messages are located in categories Conceptual Question and Debugging
Help corroborate the findings of \cite{Ghimire24} which showed that the
questions are also localized in the same categories.

Multiple Question exercise resolutions represented 14.9\% of the responses.
These prompts typically involved students submitting one or more exercises to
CharlieBot and requesting solutions. The conversation 13 in \ref{fig:graph2}
illustrates a sequence of multiple choice questions.

Requests for the bot to generate a Code Snippet or Complete Solution accounted
for 8.2\% of student messages. These prompts reflected a desire for more
direct answers; however, it was observed that, after receiving these solutions,
many students transitioned to a more active approach in seeking to understand
the code or the underlying concepts. Thus, it can be inferred that the prompts
within the Code Snippet or Complete Solution category were often used as a
starting point for a more in-depth study.

Student corrections to bot responses accounted for 4.2\% of the interactions.
In most cases, these corrections occurred when the student already had the
correct answer to an exercise and identified an error in the bot response, as
illustrated in conversation 2 in Figure \ref{fig:graph2}. These corrections
highlight that, although the bot provides quick feedback, it is not always
accurate.

Figure \ref{fig:graph1} also indicates that approximately 5.8\% of the prompts
written by students were classified as Uncategorized. These messages include
expressions of gratitude toward the bot, such as a simple '\textit{thanks}',
contextual statements like '\textit{I'm reviewing object orientation}', as well
as requests for additional exercises and summaries. Depending on the type of
investigation, there is opportunity for these messages to be further subdivided
into new categories.

Finally, 1.3\% of the messages were classified as Language Change. These prompts
were typically requests for the bot to switch languages, as seen in conversation
9 in Figure \ref{fig:graph2}.

\begin{table*}[htbp]
  \caption{Average and Standard Deviation of Different Categories (in minutes)}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{4cm} p{4cm} p{4cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
      \hline
      Code Snippet & 1.26 & 1.74 \\
      Uncategorized & 1.65 & 3.54 \\
      Language Change & 1.89 & 3.28 \\
      Multiple Question & 2.65 & 6.10 \\
      Student Correction & 3.17 & 7.33 \\
      \hdashline
      \textit{Overall} & \textit{3.43} & \textit{6.60} \\
      \hdashline
      Debugging Help & 4.84 & 10.13 \\
      Conceptual Questions & 5.41 & 10.54 \\
      Complete Solution & 6.55 & 10.14 \\
      \hline
    \end{tabular}
    \label{tab:averages}
  \end{center}
\end{table*}

@@

About the second research question: \textit{RQ2: Was there spacing between
student prompts, and if so, which category led to the most rapid and consistent
interactions with the bot?}

For the analysis of the spacing, we initially considered a minimum interval of
60 minutes as a reference to define the significant spacing \citep{Gadella24}.
Based on this, our results showed that, of the 81 conversations analyzed, only
27 (33.3\%) had at least one message with an interval greater than 60 minutes.
Figure 2 illustrates the elapsed time between messages (delta). For example,
in conversation 15, the first message was sent at time zero, while the second
message was sent by the student after a delta of 0.6 minutes. Additionally,
message 4 in conversation 15, as shown in Figure \ref{fig:graph2}, presents a
delta of 69.1 minutes, which, according to our reference, characterizes
significant spacing. Figure \ref{fig:graph3} also shows the number of spacings
in each conversation. Among conversations that exhibited spacing, the
average number of subsections per conversation was 2.2. Therefore, a
conversation with one instance of spacing is typically divided into two distinct
study subsections.

This data suggests that there is indeed little spacing between study sessions.
It is worth noting that, in order to create a new section in CharlieBot,
students must close the browser tab and start a new conversation, which results
in losing the previous chat history.

\subsubsection{Interaction Time per Category}

When a student requests a code snippet (CS), it is generally predictable that
they will resume interaction with the bot within a short period of time, as
indicated by the mean (1.26) and standard deviation (1.74) of the Code Snippet
category in Table \ref{tab:averages}. This result is noteworthy because,
although some requests are simple and lead to quick responses, others may be
more complex and would require more time for reflection. However, the low
average response time and minimal variability may indicate that students adopt a
more passive learning posture when requesting the bot to generate small pieces
of code.

Despite some occasional variation (3.54), prompts classified as Uncategorized
exhibit a relatively low average time (1.65) for students to resume interaction
with the bot. This variation occurs, in part, due to situations where students
end the conversation with an uncategorized message, such as a simple
\textit{'thanks'} but later choose to resume the interaction.

When a student requests the bot to change the language (LC) in a response, the
new interaction happens quickly (1.89) with little variation in response time
(3.28), meaning it is predictable that the student will continue interacting.

Although still below the overall average, students take a moderate amount of
time (2.65) to resume interaction with the bot after requesting the solution to
a multiple-choice question (MQ). The variation (6.10) suggests that some
responses may lead to a slower follow-up interaction, possibly indicating a
moment of deeper reflection on the part of the students. However, since both the
mean and standard deviation are below the overall average, this could indicate
a more passive behavior, as students obtain ready-made answers from the bot.

% This less active learning approach can occur for various reasons, ranging from
% the belief that reviewing solutions to similar questions enhances learning, to
% issues such as negligence, poor planning, or other possible factors.

The average time for students to resume interaction in the correction category
(SC) is close to the overall average (3.17), although the high standard
deviation (7.33) indicates significant variation. In some cases, students are
already familiar with the answer and only seek a quick correction from the bot;
in others, they detect flaws in the bot's responses and need to carefully
examine the subsequent answer to identify potential errors, which suggests a
more active behavior.

After a question classified as Debugging Help (DH), there is both a
longer-than-average response time (4.84) and a high variation in the time taken
for a new interaction (10.13). This suggests that student prompts in this
category tend to be more time-consuming and unpredictable. A possible
explanation is that some DH issues are simple, such as '\textit{what does x++ do?}',
while others involve requests for analysis of more complex code, like
\textit{'explain this piece of code (accompanied by a complete method).'}

Conceptual questions (CQ) generally result in a high average time (5.41) for
students to re-engage with the bot. Moreover, the high standard deviation
(10.54) indicates an unpredictable variation in response times. This
unpredictability may be attributed to the fact that, in some cases, the
complexity of the question and answer demands more reflection time from the
student, whereas in others, lesser complexity allows for a quicker interaction
with the bot.

Finally, a message classified as a Complete Solution request (SCO) exhibits the
longest average time (6.55) for students to resume interaction with the bot.
Additionally, the high standard deviation (10.14) highlights significant
unpredictability in response time. After requesting a complete solution,
students tend to take longer to return, due to the complexity of the problems
involved, such as the implementation of sorting algorithms like Heapsort. Table
\ref{tab:averages} presents the means and standard deviations for each
category.

However, some categories suggest quicker interactions with less reflection on
the part of students, such as 'Code Snippet' and 'Multiple Choice.' On the other
hand, other categories indicate that students may be in a more reflective
process, such as 'Debugging Help,' 'Conceptual Question,' and 'Complete
Solution.' This demonstrates that the level of student engagement varies
according to the nature of the requested interaction, reflecting different
approaches and depths in the learning process.

\subsection{Interleaving}

About the third research question: \textit{RQ3: Do students' interactions with
the bot alternate between study topics to suggest interleaving?}

To analyze the interweaving of topics, sections were initially divided into
those that exhibited spacing and those that did not.

In conversations with the bot that did not include spacing (66.7\% of the total
conversations), approximately 19.2\% of the messages resulted in a change of
topic. Qualitative analysis of what led to this interweaving indicates that
part of it was due to students' requests for assistance in solving exercises
from different topics, suggesting a less active learning approach, as the time
interval for this type of interaction tends to be shorter than average.
Additionally, some instances of interweaving were observed after a prolonged
period of 20 to 50 minutes, still below the 60-minute threshold established as
the cutoff for considering spacing.

In interactions with the bot that included spacing (33.3\% of the total
conversations), approximately 31.4\% of the messages switched topics.
Qualitative analysis of these interactions reveals that a significant portion
of this interweaving was caused by spacing. In other words, some students use
the bot for quick consultations, asking a question and then returning hours
later with another question on a different subject.

Thus, the data revealed that the existing interweaving is limited and does not
suggest an effective study strategy, as it mainly arises from requests for
exercise resolution and from the natural spacing pattern of a consultation
interaction.

@@

The objective of this work was to analyze the interaction of second-semester
Java programming students with an educational bot developed using the RAG
technique, employing two metacognitive strategies: spacing and interleaving.

Initially, the theory of metacognition, alongside the classification of student
actions, has proven effective in understanding and revealing how students
utilize the bot. Although the data shows that most messages fall within the
categories of Conceptual Questions and Debugging Help categories identified as
promoting slower but potentially deeper engagement with the bot, indicative of a
more active form of learning—the findings also reveal that students' use of
spacing and interleaving strategies remains quite limited. Encouraging students
to apply these strategies more intentionally could significantly enhance their
learning outcomes. To further explore this, a broader study is planned to
examine, for instance, the relationship between student profiles and their
strategies in using a Generative AI tool. Additionally, to support educators
in analyzing student behavior and providing more targeted assistance, an AI
model is being developed to automatically classify student messages, offering
valuable insights into the ways these tools are being employed.

% The results suggest that most student interactions with GenAI tools are centered
% on active learning strategies, although the deliberate use of interleaving and
% spacing techniques remains limited. Guiding students toward a more intentional
% application of these strategies could enhance their learning outcomes. The
% anonymity of interactions leaves it unclear whether these strategies were used
% deliberately or emerged naturally, highlighting the need for future research on
% students' conscious application of these methods. To support educators in
% analyzing student behavior and offering more targeted assistance, an AI model
% is being developed to automatically classify student messages, providing
% valuable insights into how these tools are being utilized.

@@
